---
title: "Example Data"
author: "Paolo Bosetti"
date-modified: "`{r} Sys.time()`"
date-format: "YYYYMMDD"
---

```{r setup, include=FALSE}
library(tidyverse)
library(glue)
library(modelr)
```

During the courses, example data files are used to illustrate different concepts. These files are available here for download or for direct usage.


## Loading the data files

Each data file can be separately loaded from the URLs reported [down below](#list-of-csv-files). Nonetheless, you can also exploit the ability of R `read*` functions to open a file directly from its URL. To do so, it can be useful to define a utility function like the following:

```{r}
example_url <- function(example) {
  url = paste0("https://paolobosetti.quarto.pub/data/", example)
  return(url)
}
```

Load this function in your script or put it into a utility library.
Then, just pipe it with the proper file-reading function, e.g.:

```{r eval=FALSE}
example_url("timeseries.csv") %>% read_csv()
```

or

```{r eval=FALSE}
example_url("3dprint.dat") %>% read_table(comment="#")
```



## List of CSV files

```{r echo=FALSE}
files <- list.files(path = "data", pattern = ".csv", full.names = FALSE)
reduce(files, ~paste0(.x, glue("* [{.y}](data/{.y})"), "\n"), .init="") %>% 
  knitr::asis_output()
```
## List of DAT files

```{r echo=FALSE}
files <- list.files(path = "data", pattern = ".dat", full.names = FALSE)
reduce(files, ~paste0(.x, glue("* [{.y}](data/{.y})"), "\n"), .init="") %>% 
  knitr::asis_output()
```


# Example

Let's look at the `train.csv` file from the Titanic dataset. We load the dataset and perform some basic data manipulation to only keep the relevant columns. We also classify the passengers into age classes.

```{r}
titanic <- example_url("train.csv") %>% read_csv(show_col_types = FALSE)
titanic <- titanic %>% mutate(
  Survived = as.logical(Survived),
  Pclass = as.factor(Pclass),
  AgeClass = cut(Age, breaks=c(0, 10, 18, 40, 60, Inf), labels=c("child", "young", "adult", "senior", "elderly")),
  Age = as.factor(Age)
) %>% 
  select(
    PassengerId,
    Survived,
    Pclass,
    Sex,
    Age,
    AgeClass,
    Fare
  )
titanic %>% head() %>% knitr::kable()
```

Let's plot the survival rate by age and sex class, as a histogram.

```{r}
titanic %>% 
  ggplot(aes(x=AgeClass, fill=Survived)) +
  geom_bar() +
  facet_wrap(~Sex) +
  labs(title="Survival by Age Class", x="Age Class", y="Count", fill="Survived")
```

In a more refined way, we can build a [generalized linear model of binomial type](/slides/ADAS/3-regressione.html#regressione-lineare-generalizzata)🇮🇹 to predict the survival rate based on the fare, age class, and sex:

```{r warning=FALSE}
model <- glm(Survived ~ Fare + AgeClass + Sex, data=titanic, family=binomial)
titanic <- add_predictions(titanic, model, var="pred.glm", type="response")

pred <- expand.grid(
  Fare = seq(0, 500, 5),
  AgeClass = levels(titanic$AgeClass),
  Sex = c("female", "male")
) %>% 
  add_predictions(model, var="pred.glm", type="response")

titanic %>%
  ggplot(aes(x=Fare, y=pred.glm, color=AgeClass)) +
  geom_line(data=pred) +
  geom_point(aes(shape=Survived), size=2) +
  coord_cartesian(ylim=c(0, 1)) +
  facet_wrap(~Sex) +
  scale_shape_manual(values=c(3, 1)) +
  labs(y="Predicted Survival", color="Age Class")
```
:::{.callout-tip title="A word to the wise"}
According to our analysis, if you embark the Titanic, you better be a **young and rich female** 😉.
:::