---
title: "Example Data"
date-modified: "`{r} Sys.time()`"
date-format: "YYYYMMDD"
---

```{r setup, include=FALSE}
library(tidyverse)
library(glue)
```

During the courses, example data files are used to illustrate different concepts. These files are available here for download or for direct usage.


## Loading the data files

Each data file can be separately loaded from the URLs reported [down below](#list-of-csv-files). Nonetheless, you can also exploit the ability of R `read*` functions to open a file directly from its URL. To do so, it can be useful to define a utility function like the following:

```{r eval=FALSE}
example_url <- function(example) {
  url = paste0("https://paolobosetti.quarto.pub/data/", example)
  return(url)
}
```

Load this function in your script or put it into a utility library.
Then, just pipe it with the proper file-reading function, e.g.:

```{r eval=FALSE}
example_url("timeseries.csv") %>% read_csv()
```

or

```{r eval=FALSE}
example_url("3dprint.dat") %>% read_table(comment="#")
```



## List of CSV files

```{r echo=FALSE}
files <- list.files(path = "data", pattern = ".csv", full.names = FALSE)
reduce(files, ~paste0(.x, glue("* [{.y}](data/{.y})"), "\n"), .init="") %>% 
  knitr::asis_output()
```
## List of DAT files

```{r echo=FALSE}
files <- list.files(path = "data", pattern = ".dat", full.names = FALSE)
reduce(files, ~paste0(.x, glue("* [{.y}](data/{.y})"), "\n"), .init="") %>% 
  knitr::asis_output()
```
